---
title: "Preliminary tables"
format:
  html:
    code-fold: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center", fig.retina = 3,
                      fig.width = 6, fig.height = (6 * 0.618),
                      out.width = "80%", collapse = TRUE,
                      dev = "png", dev.args = list(type = "cairo-png"))

options(digits = 3, width = 120,
        dplyr.summarise.inform = FALSE,
        knitr.kable.NA = "")
```

```{r load-libraries, warning=FALSE, message=FALSE}
library(tidyverse)
library(targets)
library(broom)
library(modelsummary)
library(kableExtra)
library(ordinal)

# Generated via random.org
set.seed(2385)

# Load data 
withr::with_dir(here::here(), {
  # Final data
  tar_load(daily_panel)
})
```

```{r}
cm <- c("derogation_ineffectTRUE" = "Derogation in effect",
        "new_cases" = "New cases",
        "cumulative_cases" = "Cumulative cases",
        "new_deaths" = "New deaths",
        "cumulative_deaths" = "Cumulative deaths",
        "year_quarter" = "Year quarter",
        "v2x_rule" = "Rule of law",
        "v2x_civlib" = "Civil liberties",
        "v2xcs_ccsi" = "Core civil society index",
        "v2csreprss" = "Civil society repression",
        "(Intercept)" = "Constant",
        "None|Minor" = "Cut: None | Minor",
        "Minor|Moderate" = "Cut: Minor | Moderate",
        "Moderate|Major" = "Cut: Moderate | Major")

gm <- tribble(
  ~raw,        ~clean,     ~fmt, ~omit,
  "nobs",      "N",        0,    FALSE,
  "nclusters", "Clusters", 0,    FALSE
)
```


# Preliminary hypothesis: Does having a derogation matter for pandemic policies?

## Table 1: ICCPR Derogations and Covid-19 Restriction Policies

```{r warning=FALSE}
model_1 <- glm(c3_cancel_events_bin ~ derogation_ineffect + new_cases + 
                 cumulative_cases + new_deaths + cumulative_deaths,
               data = daily_panel,
               family = binomial(link = "logit"))

model_2 <- glm(c4_gatherings_bin ~ derogation_ineffect + new_cases + 
                 cumulative_cases + new_deaths + cumulative_deaths,
               data = daily_panel,
               family = binomial(link = "logit"))

model_3 <- glm(c5_public_transport_bin ~ derogation_ineffect + new_cases + 
                 cumulative_cases + new_deaths + cumulative_deaths,
               data = daily_panel,
               family = binomial(link = "logit"))

model_4 <- glm(c7_internal_movement_bin ~ derogation_ineffect + new_cases + 
                 cumulative_cases + new_deaths + cumulative_deaths,
               data = daily_panel,
               family = binomial(link = "logit"))

model_5 <- glm(c8_intl_travel_bin ~ derogation_ineffect + new_cases + 
                 cumulative_cases + new_deaths + cumulative_deaths,
               data = daily_panel,
               family = binomial(link = "logit"))

modelsummary(list("Cancel Public Event" = model_1,
                  "Gathering Restrictions" = model_2,
                  "Close Public Transit" = model_3,
                  "Movement"  = model_4,
                  "International Travel" = model_5),
             vcov = ~country_name, 
             fmt = 4, stars = c("*" = 0.1, "**" = 0.01, "***" = 0.001),
             coef_map = cm, gof_map = gm)
```

## Table 2: ICCPR Derogations and Covid-19 Restriction Policies with more controls

```{r warning=FALSE}
model_6 <- glm(c3_cancel_events_bin ~ derogation_ineffect + new_cases + 
                 cumulative_cases + new_deaths + cumulative_deaths +
                 year_quarter + v2x_rule + v2x_civlib + v2xcs_ccsi,
               data = daily_panel,
               family = binomial(link = "logit"))

model_7 <- glm(c4_gatherings_bin ~ derogation_ineffect + new_cases + 
                 cumulative_cases + new_deaths + cumulative_deaths +
                 year_quarter + v2x_rule + v2x_civlib + v2xcs_ccsi,
               data = daily_panel,
               family = binomial(link = "logit"))

model_8 <- glm(c5_public_transport_bin ~ derogation_ineffect + new_cases + 
                 cumulative_cases + new_deaths + cumulative_deaths +
                 year_quarter + v2x_rule + v2x_civlib + v2xcs_ccsi,
               data = daily_panel,
               family = binomial(link = "logit"))

model_9 <- glm(c7_internal_movement_bin ~ derogation_ineffect + new_cases + 
                 cumulative_cases + new_deaths + cumulative_deaths +
                 year_quarter + v2x_rule + v2x_civlib + v2xcs_ccsi,
               data = daily_panel,
               family = binomial(link = "logit"))

model_10 <- glm(c8_intl_travel_bin ~ derogation_ineffect + new_cases + 
                 cumulative_cases + new_deaths + cumulative_deaths +
                  year_quarter + v2x_rule + v2x_civlib + v2xcs_ccsi,
               data = daily_panel,
               family = binomial(link = "logit"))

modelsummary(list("Cancel Public Event" = model_6,
                  "Gathering Restrictions" = model_7,
                  "Close Public Transit" = model_8,
                  "Movement"  = model_9,
                  "International Travel" = model_10),
             vcov = ~country_name, 
             fmt = 4, stars = c("*" = 0.1, "**" = 0.01, "***" = 0.001),
             coef_map = cm, gof_map = gm)
```


# H~1a-1c~

## Table 3: ICCPR Derogations and Covid-19 Polices

```{r}
model_11 <- clm(pandem_discrim ~ derogation_ineffect + new_cases + 
                  cumulative_cases + new_deaths + cumulative_deaths +
                  year_quarter + v2x_rule + v2x_civlib + v2xcs_ccsi,
                data = daily_panel)

model_12 <- glm(pandem_ndrights ~ derogation_ineffect + new_cases + 
                   cumulative_cases + new_deaths + cumulative_deaths +
                   year_quarter + v2x_rule + v2x_civlib + v2xcs_ccsi,
                 data = daily_panel,
                 family = binomial(link = "logit"))

model_13 <- clm(pandem_nolimit ~ derogation_ineffect + new_cases + 
                   cumulative_cases + new_deaths + cumulative_deaths +
                   year_quarter + v2x_rule + v2x_civlib + v2xcs_ccsi,
                 data = daily_panel)

model_14 <- clm(pandem_abusive ~ derogation_ineffect + new_cases + 
                   cumulative_cases + new_deaths + cumulative_deaths +
                   year_quarter + v2x_rule + v2x_civlib + v2xcs_ccsi,
                 data = daily_panel)

modelsummary(list("Discriminatory Policy" = model_11,
                  "Non-Derogable Rights" = model_12,
                  "No Time Limit Measures" = model_13,
                  "Abusive Enforcement"  = model_14),
             vcov = ~country_name, 
             fmt = 4, stars = c("*" = 0.1, "**" = 0.01, "***" = 0.001),
             coef_map = cm, gof_map = gm)
```

# H~2a-2d~

## Table 4: Repressive Towards Civil Society and Covid-19 Policies

```{r}
model_15 <- clm(pandem_discrim ~ derogation_ineffect + new_cases + 
                  cumulative_cases + new_deaths + cumulative_deaths +
                  year_quarter + v2x_rule + v2x_civlib + v2xcs_ccsi + v2csreprss,
                data = daily_panel)

model_16 <- glm(pandem_ndrights ~ derogation_ineffect + new_cases + 
                   cumulative_cases + new_deaths + cumulative_deaths +
                   year_quarter + v2x_rule + v2x_civlib + v2xcs_ccsi + v2csreprss,
                 data = daily_panel,
                 family = binomial(link = "logit"))

model_17 <- clm(pandem_media ~ derogation_ineffect + new_cases + 
                   cumulative_cases + new_deaths + cumulative_deaths +
                   year_quarter + v2x_rule + v2x_civlib + v2xcs_ccsi + v2csreprss,
                 data = daily_panel)

model_18 <- clm(pandem_abusive ~ derogation_ineffect + new_cases + 
                   cumulative_cases + new_deaths + cumulative_deaths +
                   year_quarter + v2x_rule + v2x_civlib + v2xcs_ccsi + v2csreprss,
                 data = daily_panel)

modelsummary(list("Discriminatory Policy" = model_15,
                  "Non-Derogable Rights" = model_16,
                  "Limiting media" = model_17,
                  "Abusive Enforcement"  = model_18),
             vcov = ~country_name, 
             fmt = 4, stars = c("*" = 0.1, "**" = 0.01, "***" = 0.001),
             coef_map = cm, gof_map = gm)
```

## Table 5: ICCPR Derogations and Supportive Economic Policies

```{r}
model_19 <- glm(e1_income_support_bin ~ derogation_ineffect + new_cases + 
                   cumulative_cases + new_deaths + cumulative_deaths +
                   year_quarter + v2x_rule + v2x_civlib + v2xcs_ccsi,
                 data = daily_panel,
                 family = binomial(link = "logit"))

model_20 <- glm(e2_debt_relief_bin ~ derogation_ineffect + new_cases + 
                   cumulative_cases + new_deaths + cumulative_deaths +
                   year_quarter + v2x_rule + v2x_civlib + v2xcs_ccsi,
                 data = daily_panel,
                 family = binomial(link = "logit"))

modelsummary(list("Income Support" = model_19,
                  "Debt Relief" = model_20),
             vcov = ~country_name, 
             fmt = 4, stars = c("*" = 0.1, "**" = 0.01, "***" = 0.001),
             coef_map = cm, gof_map = gm)
```
